\vspace{-1cm} Model complexity breeds contempt: using non-linear models to compare basic properties of species' distributions

### 2021-02-27 manuscript/manuscript.tex
Finished methods section

%Title paper
\title{\vspace{-1cm}
Model simplicity breeds contempt: simple models to compare properties of species' distributions}
\author[1,*]{\normalsize Bernat Bramon Mora}
\author[1]{\normalsize Jake M.\ Alexander}
\affil[1]{\footnotesize Institute of Integrative Biology, ETH Zürich, Zürich, Switzerland}
\affil[*]{\footnotesize  bernat.bramon@gmail.com}

@@

Many ecological textbooks \citep{krebsEcologyExperimentalAnalysis1972} assume the shape of species distributions to be unimodal and symmetric, but some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austin2002spatial}. In practice, distribution frameworks often use logistic regressions with a linear relationship between covariates (but see XX and YY). This is useful because it simplifies the optimization process, but it comes with several statistical shortcomings. First and foremost, such response curve and the linear relationship between covariates often comes with a set of implicit mathematical constrains that might not be biologically justified. From a purely statistical perspective, if all that we are willing to assume is that species occupy finite geographic ranges---i.e.~their probability distributions have finite variance---the most conservative statistical approach is to model these as a Gaussian distributions \citep{frankCommonPatternsNature2009}. This is rarely the starting point in most statistical frameworks that study general biodiversity patterns (but see ref), choosing to use instead Gaussian-logit response curves (refs). Other factors might then condition species distributions to showcase heavy-tails or a skewed shapes, revealing interesting ecological processes shaping biodiversity patterns \citep{austinNonlinearSpeciesResponse1976, minchinEvaluationRelativeRobustness1987}. The starting point, nevertheless, should be the one that makes the fewest assumptions (i.e.~the maximum entropy distribution \citealt{entropy stuff}), and every new shape will imply a hypotheses on how communities are distributed \citep{damenSpatialPredictionsCommunity2017}. Second, the aforementioned structural constrains also limit our ability to include any prior information to our parameter estimates. Observations on species' geographic variation and optimal climatic conditions have long been documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge on species' realized niches (e.g.~\citealt{landoltFloraIndicativaOkologische2010}). That said, this information is rarely accounted for in most modelling approaches, mainly because there is not a straightforward way to feed this information into the parameters of a linear model (\citealt{scherrerEcologicalIndicatorValues2019}; but see \citealt{terbraakWeightedAveragingLogistic1986}). Finally, and perhaps most importantly, a direct biological interpretation of parameter estimates in linear models becomes increasingly difficult as one moves from unimodal and symmetric distributions \citep{jamilGeneralizedLinearMixed2013, terbraakWeightedAveragingLogistic1986} to skewed distributions \citep{huismanHierarchicalSetModels1993}, making the tests of hypothesis on global biodiversity patterns particularly challenging. For example, \citet{huismanHierarchicalSetModels1993} proposed several non-linear models to characterize several features of species' response curves; however, species' environmental indicator values, range size or distribution skewness are difficult to understand altogether following these model structures.

@@

The field of ecology has quickly moved towards mechanistic and process-based approaches to understand species' distributions \citep{wartonManyVariablesJoint2015}. This has resulted in a plethora of models accounting for several biotic and abiotic factors into the predictions of species co-occurrence. Here, we instead rethink traditional modelling approaches and develop a conceptually simple---and yet statistical and computationally complex---statistical framework to revisit some classic hypothesis in ecology and biogeoraphy. In particular, we develop a Bayesian hierarchical model that accounts for all prior information that we have regarding the distribution of alpine plant species along an elevation gradient in the Swiss Alps, including expert knowledge on species environmental indicator values, range sizes, and plant physiology. We start by considering species' response curves as Gaussian distributed, and then we adapt our model to allow for skewed and long-tailed distributions. Using this statistical framework, we are able to compare the basic properties of the realized niches of multiple species, testing for the existence of general biogeographical patterns. First, we test for the Rapopor's rule, which predicts a positive relationship between range size and elevation \citep{stevensElevationalGradientAltitudinal1992}. While this pattern has been largely studied for multiple systems and across gradients \citep{mccainElevationalRapoportRule2013}; contrasting evidence suggests this rule not to be pervasive across species \citep{ribasRapoportEffectWidespread2006, bhattaraiCanRapoportRule2006, mccainElevationalRapoportRule2013}. Our results not only allow us to properly test the existence of this geographical pattern, but they also showcase variation in how different types of species, such as native or neophytes, might respond to an environmental gradient. Second, we study whether or not species' distributions show steeper declines towards stressful conditions, testing the so-called abiotic stress limitation hypothesis (ref). \citet{normandImportanceAbioticStress2009} tested this for vegetation data using \citeauthor{huismanHierarchicalSetModels1993}'s statistical models for several independent species, finding no clear support for such a hypothesis. Our results are able to shed light on this geographical pattern as well as to highlight the degree to which different species will showcase different levels of decline towards stressful conditions. Specifically, we are able to link plant physiological traits to the skewness of their distributions. Overall, we use models that are solely constrained by the empirical information that we truly have regarding our system, relaxing as much as possible the structural constrains of the statistical framework. Using these models, we are able uncover the true shape of empirical plant distributions and answer fundamental questions regarding the way systems of many species are distributed along environmental gradients.

@@

\subsubsection*{Floristic data}
To complement the aforementioned distribution data, we used a floristic database of most vegetation across Switzerland. This database was build based on expert knowledge and field experience of botanists and ecologists, and contains information regarding species' environmental preferences and physiological traits. Species' environmental preferences in this database can be used to inform distribution models---e.g.~as an informative prior in a Bayesian framework. These are characterized following the ecological indicator values developed by \citet{landoltFloraIndicativaOkologische2010}, providing both an estimate of the average conditions in which a species can be found as well as a broad description of their range of variation. These values are provided for a range of 10 climatic variables, including temperature, continentality, light conditions, as well as moisture, acidity and nutrient content of the soil (see a full list and description of the ecological indicators in the Supplementary Methods; \citealt{landoltFloraIndicativaOkologische2010}). On the other hand, the information regarding species' physiological traits represents general descriptions of species' growth and life strategies---examples include their growth forms, nature of the storage organs, dispersal ability and pollinator agents. In total, we identify more than $120$ binary traits that characterize the physiology of species (see a full list and description of the ecological indicators in the Supplementary Methods; \citealt{landoltFloraIndicativaOkologische2010}).

@@

\subsubsection*{Sampling the posterior}
We generated the posterior samples for the Bayesian models with the help of the R package `rstan' to \citep{standevelopentteamRStanInterfaceStan2021}. Sampling models like the ones described above can be computationally very expensive. This is especially true when using ordered categorical likelihood functions (see \citealt{standevelopmentteamStanModelingLanguage2021}). Therefore, we focus on those species for which we have more than 30 occurrences when modelling ordinal data, which is the case for the majority of the results of this work. When using presence/absence data, we limit our study to those species for which have more than 10 occurrences.

To test the performance of the model as well as our choice of prior distributions, we modelled simulated data and compared the sampled posterior distributions to the data-generating parameters (see Supplementary Methods; Supplementary Fig. 2). Notice that using the link function in Eq. \ref{eq:baseline-response} could cause problems when sampling the model, and some adjustments need to be made when specifying the model (see Supplementary Methods and the Code Availability section).

\subsection*{Modifying the baseline model}
We proposed a baseline model that is naive in terms of the structural assumptions regarding the data, and yet accounts for all information we truly have about the system. Modifying this model, we can now test hypotheses regarding the properties of empirical species' distributions. To propose new species' response curves, however, we want to ensure two key conditions: (i) the probability distribution must have defined variance, and (ii) the Gaussian shape must be a special case of the probability distribution.

\subsubsection*{Heavy-tail response curve}

\subsubsection*{Skewed response curve}

\subsubsection*{Heavy-tail and skewed response curve}

\subsection*{Alternative variance-covariance structures}
The model structure defined in Eq.~(\ref{eq:baseline}) allows us to test the effect of adding new information. Specifically, we can do this by modifying Eq.~(\ref{eq:covariance}). For example, imagine that we have multiple matrices $D^k$ characterizing species' differences along different axis of variation---i.e.~two matrices characterizing ecological and environmental traits, or multiple matrices resulting from the different group memberships estimated using the MMSBM. One could modify Eq.~(\ref{eq:covariance}) for a particular parameter---e.g.~parameter $\alpha_i$---such that
\begin{equation} 
\Sigma^{\alpha}_{ij} = \eta_{\alpha}\,\text{exp}\left(-\sum_k\rho_{\alpha k} {D^{k}_{ij}}^2\right) + \delta_{ij} \sigma_{\alpha} ,
\label{eq:covariance-complex}
\end{equation}
where now $\rho_{\alpha k}$ are separate relevance hyperparameters for each distance matrix in the total variance of $\alpha_i$. Notice that the same is true for the covariance of parameters $\beta_i^k$ and $\lambda_i^k$. Finally, for all hyperparameters and as described for the baseline model, we use adaptive priors across covariance structures.

Given $y_{ij}$ the presence/absence of any species $i$ in any given site $j$, and a set of $k$ environmental variables $x_{jk}$, we estimate species' distributions as:
\begin{equation} 
\begin{split}
y_{ij} & \sim \text{Binomial}\left(1, p_{ij}\right)\\
\text{log}\left(p_{ij}\right) & = -\alpha_{i} - \sum_{k} \lambda_{ik} \left(x_{jk}-\beta_{ik}\right)^2\\
\text{log}(\alpha)  & \sim \text{MVNormal}\Big(\hat{\alpha}, \Sigma^{\alpha}\Big)\\
\beta_{ik}  & \sim \text{MVNormal}\left(\hat{\beta}_{k}, \Sigma^{\beta_{k}}\right)\\
\text{log}(\lambda_{ik})  & \sim \text{MVNormal}\left(\hat{\lambda_{k}}, \Sigma^{\lambda_{k}}\right)\\
\hat{\alpha}, 
\hat{\lambda^{k}}, 
\hat{\beta^{k}}  & \sim \text{Normal}\left(0,1\right)
\end{split}
\label{eq:baseline}
\end{equation}
Notice that this model structure assumes all plants to have a uni-modal distributions along each environmental axis (see the model's behaviour in Supplementary Figure XX), where parameters $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$ describe amplitude of the probability $p_{ij}$, species' average climatic suitability and range of variation along the different environmental gradients, respectively\footnote[2]{I'll rewrite the likelihood function to an ordered categorical as soon as I get things to work properly with count data.}. While potentially sacrificing predictive accuracy, this model structure allows us to explicitly incorporate all prior knowledge that we have regarding species' distributions via $\Sigma^{\alpha}$, $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$. More specifically, we express $\beta_i^k$ and $\log\left(\lambda_i^k\right)$ as multivariate normal distributions---i.e.~Gaussian processes---such that $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$ are variance-covariance matrices describing species' similarity in terms of their average climatic suitability and range of variation along the different environmental gradients, respectively. Likewise, $\log\left(\alpha\right)$ is characterized as a Gaussian Process, where the corresponding variance-covariance matrix $\Sigma^{\alpha}$ is designed to also incorporate some of the prior information that we have with regards to species' physiological traits.

In all cases, all variance-covariance matrices are defined as follows:
\begin{equation} 
\Sigma^{\chi}_{ij} = \eta_{\chi}\,\text{exp}\left(-\rho_{\chi} {D^{\chi}_{ij}}^2\right) + \delta_{ij} \sigma_{\chi} ,
\label{eq:covariance}
\end{equation}

where $\Sigma^{\chi}_{ij}$ describes the covariance between any pair of species $i$ and $j$ for any given parameter $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$. Following this expression, such covariance declines exponentially with the square of the different $D^{\chi}_{ij}$, which are distance measures computed using the prior information that we have regarding species' distributions. Specifically, given $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$, the distance measures are calculated using plants' physiological traits, ecological indicator values and range of variation, respectively (see below for further details). For each covariance matrix, the hyperparameter $\rho_{\chi}$ determines the rate of decline of the covariance between any two species, and $\eta_{\chi}$ defines its maximum value. The hyperparameter $\sigma_{\chi}$ describes the additional covariance between the different observations for any given species. For any given hyperparameter, we choose adaptive priors across covariance structures. That is, and taking $\rho_{\chi}$ as an example, we choose a prior $\log\left(\rho_{\chi}\right)\sim \text{Normal}\left(\hat{\rho}, \sigma_{\rho}\right)$ such that $\hat{\rho}\sim \text{Normal}\left(0, 1\right)$ and $\sigma_{\rho}\sim \text{Exponential}\left(1\right)$. Similar priors were chosen for both $\eta_{\chi}$ and $\sigma_{\chi}$. We generated the posterior samples for the Bayesian models with the help of the R package `rstan' to \citep{rstan}.

\subsection*{Covariance matrices from incomplete categorical and ordinal data}
The prior information that we have regarding species' distributions is represented by the set of ordinal and categorical traits found in the floristic database. More specifically, both the ecological indicator values and range of variation are ordinal traits specified for all species, whereas plants' physiological data are characterized by categorical data containing multiple missing entries. These data could be directly used as covariates in any given distribution model; however, we want this information to be accounted for as a prior for the parameters of our Bayesian model. To do so, we need to compile the traits in the floristic database into variance-covariance matrices characterizing the \textit{a priori} similarity between species.

We define these variance-covariance matrices are defined as follows:
\begin{equation} 
\Sigma_{ij} = \eta\,\text{exp}\left(-\rho {D_{ij}}^2\right) + \delta_{ij} \sigma ,
\label{eq:covariance}
\end{equation}

where $\Sigma_{ij}$ describes the covariance between any pair of species $i$ and $j$. Following this expression, such covariance declines exponentially with the square of the different $D_{ij}$, which are distance measures computed using the prior information that we have regarding species' distributions. Specifically, given $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$, the distance measures are calculated using plants' physiological traits, ecological indicator values and range of variation, respectively (see below for further details). For each covariance matrix, the hyperparameter $\rho_{\chi}$ determines the rate of decline of the covariance between any two species, and $\eta_{\chi}$ defines its maximum value. The hyperparameter $\sigma_{\chi}$ describes the additional covariance between the different observations for any given species. For any given hyperparameter, we choose adaptive priors across covariance structures. That is, and taking $\rho_{\chi}$ as an example, we choose a prior $\log\left(\rho_{\chi}\right)\sim \text{Normal}\left(\hat{\rho}, \sigma_{\rho}\right)$ such that $\hat{\rho}\sim \text{Normal}\left(0, 1\right)$ and $\sigma_{\rho}\sim \text{Exponential}\left(1\right)$. Similar priors were chosen for both $\eta_{\chi}$ and $\sigma_{\chi}$. We generated the posterior samples for the Bayesian models with the help of the R package `rstan' to \citep{rstan}.

---which define the prior information that we have for $\beta_i^k$, and $\lambda_i^k$, respectively---are ordinal traits specified for all species. In contrast, the plants' physiological data---shaping the prior for the parameters $\alpha_i$---are characterized by categorical data containing multiple missing entries. Therefore, we need to carefully compile this data into distance matrices in order to be able to feed this prior information into the model.

The missing component in the description of model (\ref{eq:baseline}) is the distance matrices $D^{\chi}$ used to define the covariance matrices $\Sigma^{\alpha}$, $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$. In this model, such distance matrices characterize differences between plant species. In the floristic data, however, the prior information that we have for these differences is represented by a set of ordinal and categorical traits. More specifically, both the ecological indicator values and range of variation---which define the prior information that we have for $\beta_i^k$, and $\lambda_i^k$, respectively---are ordinal traits specified for all species. In contrast, the plants' physiological data---shaping the prior for the parameters $\alpha_i$---are characterized by categorical data containing multiple missing entries. Therefore, we need to carefully compile this data into distance matrices in order to be able to feed this prior information into the model.

More generally, we want to understand the way $N$ species are characterized by $M$ categorical traits. One way to frame this problem is by using a network representation. Following the ideas presented by \citet{godoy-loriteAccurateScalableSocial2016}, we assume that species can be connected to each of these traits by an interaction $\left(i, j\right)$ that can be of any type $r\in R$. Notice that this provides as with multiple ways to account for the information---and lack thereof---contained in the different categorical and ordinal traits $M$. That is, the $R$ types of interactions can represent the lack of information for a particular link $\left(i, j\right)$, the absence or presence of such interaction, and any type of association between $i$ and $j$.

>>>>

%Title paper
\title{\vspace{-1cm}
Model complexity breeds contempt: using non-linear models to compare basic properties of species' distributions}
% Model familiarity breeds contempt: simple models to compare properties of species' distributions
% Model simplicity breeds contempt: simple models to compare properties of species' distributions
\author[1,*]{\normalsize Bernat Bramon Mora}
\author[1]{\normalsize Jake M.\ Alexander}
\affil[1]{\footnotesize Institute of Integrative Biology, ETH Zürich, Zürich, Switzerland}
\affil[*]{\footnotesize  bernat.bramon@gmail.com}

@@

Many ecological textbooks \citep{krebsEcologyExperimentalAnalysis1972} assume the shape of species distributions to be unimodal and symmetric, but some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austin2002spatial}. In practice, distribution frameworks often use logistic regressions with a linear relationship between covariates (but see XX and YY). This is useful because it simplifies the optimization process, but it comes with some statistical shortcomings. First and foremost, such response curve and the linear relationship between covariates often comes with a set of implicit mathematical constrains that might not be biologically justified. From a purely statistical perspective, if all that we are willing to assume is that species occupy finite geographic ranges---i.e.~their probability distributions have finite variance---the most conservative statistical approach is to model these as a Gaussian distributions \citep{frankCommonPatternsNature2009}. This is rarely the starting point in most statistical frameworks that study general biodiversity patterns (but see ref), choosing to use instead Gaussian-logit response curves (refs). Other factors might then condition species distributions to showcase fat-tails or a skewed shapes, revealing interesting ecological processes shaping biodiversity patterns \citep{austinNonlinearSpeciesResponse1976, minchinEvaluationRelativeRobustness1987}. The starting point, nevertheless, should be the one that makes the fewest assumptions (i.e.~the maximum entropy distribution; \citealt{frankCommonPatternsNature2009}), and every new shape will imply a hypotheses on how communities are distributed \citep{damenSpatialPredictionsCommunity2017}. Second, the aforementioned structural constrains also limit our ability to include any prior information to our parameter estimates. Observations on species' geographic variation and optimal climatic conditions have long been documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge on species' realized niches (e.g.~\citealt{landoltFloraIndicativaOkologische2010}). That said, this information is rarely accounted for in most modelling approaches, potentially because there is not a straightforward way to feed this information into the parameters of a linear model (\citealt{scherrerEcologicalIndicatorValues2019}; but see \citealt{terbraakWeightedAveragingLogistic1986}). Finally, and perhaps most importantly, a direct biological interpretation of parameter estimates in linear models becomes increasingly difficult as one moves from unimodal and symmetric distributions \citep{terbraakWeightedAveragingLogistic1986, jamilGeneralizedLinearMixed2013} to skewed distributions \citep{huismanHierarchicalSetModels1993}, making the tests of hypothesis on global biodiversity patterns particularly challenging. For example, \citet{huismanHierarchicalSetModels1993} proposed several non-linear models to characterize several features of species' response curves; however, species' environmental indicator values, range size or distribution skewness are difficult to understand altogether following these model structures.

@@

The field of ecology has quickly moved towards mechanistic and process-based approaches to understand species' distributions \citep{wartonManyVariablesJoint2015}. This has resulted in a plethora of models accounting for several biotic and abiotic factors into the predictions of species co-occurrence. Here, we instead rethink traditional modelling approaches and develop a conceptually simple---and yet statistical and computationally complex---statistical framework to revisit some classic hypothesis in ecology and biogeoraphy. In particular, we develop a Bayesian hierarchical model that accounts for all prior information that we have regarding the distribution of alpine plant species along an elevation gradient in the Swiss Alps, including expert knowledge on species environmental indicator values, range sizes, and plant physiology. We start by considering species' response curves as Gaussian distributed, and then we adapt our model to allow for skewed and long-tailed distributions. Using this statistical framework, we are able to compare the basic properties of the realized niches of multiple species, testing for the existence of general biogeographical patterns. First, we test for the Rapopor's rule, which predicts a positive relationship between range size and elevation \citep{stevensElevationalGradientAltitudinal1992}. While this pattern has been largely studied for multiple systems and across gradients \citep{mccainElevationalRapoportRule2013}; contrasting evidence suggests this rule not to be pervasive across species \citep{ribasRapoportEffectWidespread2006, bhattaraiCanRapoportRule2006, mccainElevationalRapoportRule2013}. Our results not only allow us to properly test the existence of this geographical pattern, but they also showcase variation in how different types of species, such as native or neophytes, might respond to an environmental gradient. Second, we study whether or not species' distributions show steeper declines towards stressful conditions, testing the so-called abiotic stress limitation hypothesis (ref). \citet{normandImportanceAbioticStress2009} tested this for vegetation data using \citeauthor{huismanHierarchicalSetModels1993}'s statistical models for several independent species, finding no clear support for such a hypothesis. Our results are able to shed light on this geographical pattern as well as to highlight the degree to which different species will showcase different levels of decline towards stressful conditions. Specifically, we are able to link plant physiological traits to the skewness of their distributions. Overall, we use models that are solely constrained by the empirical information that we truly have regarding our system, relaxing as much as possible the structural constrains of the statistical framework. Using these models, we are able uncover the approximate shape of empirical plant distributions and answer fundamental questions regarding the way systems of many species are distributed along environmental gradients.

@@

\subsubsection*{Floristic data}
To complement the aforementioned distribution data, we used a floristic database of most vegetation across Switzerland. This database was build based on expert knowledge and field experience of botanists and ecologists, and contains information regarding species' environmental preferences and physiological traits. Species' environmental preferences in this database can be used to inform distribution models---e.g.~as an informative prior in a Bayesian framework. These are characterized following the ecological indicator values developed by \citet{landoltFloraIndicativaOkologische2010}, providing both an estimate of the average conditions in which a species can be found as well as a broad description of their range of variation. These values are provided for a range of 10 climatic variables, including temperature, continentality, light conditions, as well as moisture, acidity and nutrient content of the soil (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}). On the other hand, the information regarding species' physiological traits represents general descriptions of species' growth and life strategies---examples include their growth forms, nature of the storage organs, dispersal ability and pollinator agents. In total, we identify more than $120$ binary traits that characterize the physiology of species (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}). Finally, and in addition to species' environmental preferences and physiological traits, the floristic data also contains information on species types (e.g.~identifying those species that are neophytes)---and growth tendency (e.g.~indicating species that have shown decline or increase in their populations over the recent years). We describe this information in more detail in the Supplementary Table 1.

@@

\subsubsection*{Alternative variance-covariance structures}
The model structure defined above allows us to test how different sources of information characterize each of the different parameters. Specifically, we can do this by modifying Eq.~(\ref{eq:covariance-baseline}). For example, imagine that we have multiple matrices $D^k$ characterizing species' differences along different axis of variation---e.g.~two matrices characterizing physiological and environmental traits. One can modify Eq.~(\ref{eq:covariance-baseline}) for a particular parameter---e.g.~$\beta_{i}$---such that
\begin{equation} 
\Sigma_{ij} = \eta\,\text{exp}\left(-\sum_k\rho_{k} {D^{k}_{ij}}^2\right) + \delta_{ij} \sigma ,
\label{eq:covariance-complex}
\end{equation}
where now $\rho_{k}$ are separate relevance hyperparameters for each distance matrix in the total variance of $\beta_i$.

\subsubsection*{Sampling the posterior}
We generated the posterior samples for the Bayesian models with the Hamiltonian Monte Carlo algorithm implementation provided by the R package `rstan' to \citep{standevelopentteamRStanInterfaceStan2021}. Sampling models like the ones described above can be computationally very expensive. This is especially true when using ordered categorical likelihood functions (see \citealt{standevelopmentteamStanModelingLanguage2021}). Therefore, we focus on those species for which we have more than 30 occurrences when modelling ordinal data, which is the case for the majority of the results of this work. When using presence/absence data, we limit our study to those species for which have more than 10 occurrences.

To test the performance of the model as well as our choice of prior distributions, we modelled simulated data and compared the sampled posterior distributions to the data-generating parameters (see Supplementary Methods; Supplementary Fig. 2). Notice that using the link function in Eq.~(\ref{eq:baseline-response}) could cause problems when sampling the model, and some adjustments need to be made when specifying the model (see Supplementary Methods and the Code Availability section).

\subsection*{Modifying the baseline model}
We proposed a baseline model that is naive regarding how the data is distributed, and yet accounts for all prior information that we have about the system. Now, we want to modify this model to test the extent to which empirical species' distributions showcase different properties, while preserving both the interpretation of the parameter estimates and the prior information. More specifically, to propose new species' response curves, we follow three key criteria: (i) the probability distribution must have a defined variance and mean, (ii) the Gaussian shape must be a special case of the probability distribution, and (iii) there must be a re-parametrization of the model that allows us to keep the same prior information and interpretable parameters.

\subsubsection*{Fat-tail response curve}
Distributions with fat-tails are very common across fields, as they can capture processes such as seasonality (e.g.~in communications patterns; \citealt{malmgrenPoissonianExplanationHeavy2008}) or some stochastic events (e.g.~in the spread of infectious diseases; \citealt{wongEvidenceThatCoronavirus2020}). Indeed, fat-tail distributions are pervasive in ecology; for example, species' dispersal patterns have been shown to have fat-tails due to natural variability among individuals \citep{petrovskiiDispersalStatisticallyStructured2009}. Therefore, one might expect these properties to also emerge in alpine communities, where seasonality and dispersal patterns are crucial factors determining species' distributions \citep{ref}. To accommodate this feature into our baseline model, we could consider response curves that follow a generalized error or a non-standardized Student's t-distributions. In both cases, the normal distribution is a special case of them, and both contain parameters that regulate the kurtosis of the probability distribution (Supplementary Figure XX).

Starting with the generalized error distribution, we can adapt Eq.~(\ref{eq:baseline-response}) to present this non-linear form as follows:
\begin{equation}
\begin{split}
\text{log}\left(p_{ij}\right) = -\alpha_{i} - \gamma^{\prime}_{i}\, |x_{j}-\beta_{i}|^{\nu_{i}} ,
\end{split}
\label{eq:generalized-response}
\end{equation}
where $\gamma^{\prime}_{i} = g(\gamma_{i}, \nu_{i})$, and $\nu_{i}$ is a parameter that describes the kurtosis of the distribution, which we define as $\nu_{i}\in\left(1, \infty\right)$. Following this, we choose an adaptive prior for this set of new parameter such that $\log\left(\nu_{i}-1\right)\sim \text{Normal}\left(\hat{\nu}, \sigma_{\nu}\right)$, where $\hat{\nu}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\nu}\sim\text{Exponential}\left(2\right)$. Given the relationship between $\gamma^{\prime}_{i}$ and $\gamma_{i}$,  we can re-parametrize the model and follow Eq.~(\ref{eq:baseline-priors}) to define the corresponding prior distributions (see Supplementary Table 2; Code Availability section; \citealt{stats book}). Notice that the Gaussian distribution will naturally emerge when $\nu_i=2$.

In a similar fashion, we can use a non-standardized Student's t-distribution by modifying Eq.~(\ref{eq:baseline-response}). In this case, the response curve can be characterized as follows:
\begin{equation}
\begin{split}
p_{ij}  =  \alpha^{\prime}_{i}\left(1+\frac{\gamma^{\prime}_{i}}{\nu_{i}}\, \left(x_{j}-\beta_{i}\right)^2\right)^{-\frac{\nu_{i}+1}{2}},
\end{split}
\label{eq:student-response}
\end{equation}
where $\alpha^{\prime}_{i} = f_{1}\left(\alpha_{i}, \gamma^{\prime}_{i}, \nu_{i}\right)$, $\gamma^{\prime}_{i} = f_{2}\left(\gamma_{i}, \nu_{i}\right)$, and $\nu_{i}$ is a parameter that describes the kurtosis of the distribution, which we define as $\nu_{i}\in\left(2, \infty\right)$. These boundary conditions are necessary in order for the distributions to have finite variance. Similar to the description above, we choose an adaptive prior such that $\log\left(\nu_{i}-2\right)\sim \text{Normal}\left(\hat{\nu}, \sigma_{\nu}\right)$, where $\hat{\nu}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\nu}\sim\text{Exponential}\left(1\right)$. Again, we can re-parametrize the model following $f_1$ and $f_2$, and set the rest of the prior distributions as above (see Supplementary Table 2; Code Availability section). In this case, the Gaussian distribution is a special case of Eq.~(\ref{eq:student-response}) when $\nu_i\to\infty$ \citep{stats book}.

\subsubsection*{Skewed response curve}
When species experience abiotic or biotic pressures that increase or decrease along an environmental gradient, one might expect their distributions to be skewed in one direction. Likewise, this same skewed shape can also be the result of asymmetric environmental tolerance. One way to accommodate this feature to our models is by considering a skewed normal distribution (Supplementary Figure XX). As for the two cases described above, the Gaussian is a special case of this distribution, and it contains a parameter that controls for the level and direction of `skewness'. Importantly, this distribution presents normal-like tails; therefore, the added skewness does not make additional assumptions regarding how species' are distribution along the gradient. To test for the existence of this feature, we modified the species' response curve in Eq.~(\ref{eq:baseline-response}) as
\begin{equation}
\begin{split}
p_{ij} & =\hat{p}_{ij} \left[1+\text{erf}\left(\lambda_{i}\left(x_{j}-\beta^{\prime}_{i}\right)\sqrt{\frac{\gamma^{\prime}_{i}}{2}}\right)\right] \\
\text{log}\left(\hat{p}_{ij}\right) & = -\alpha^{\prime}_{i} - \gamma^{\prime}_{i} \left(x_{j}-\beta^{\prime}_{i}\right)^2 ,
\end{split}
\label{eq:skewed-response}
\end{equation}
where $\alpha^{\prime}_{i} = q_{1}\left(\alpha_{i}, \gamma^{\prime}_{i}, \lambda_{i}\right)$, $\beta^{\prime}_{i} = q_{2}\left(\gamma^{\prime}_{i}, \lambda_{i}\right)$, $\gamma^{\prime}_{i} = q_{3}\left(\gamma_{i}, \lambda_{i}\right)$, and $\lambda_{i}$ is a parameter that describes the skewness of the distribution, which we define as $\lambda_{i}\in\left(-10, 10\right)$. Notice that these boundary conditions for $\lambda_{i}$ are arbitrarily set for computational purposes, as this parameter is theoretically not bounded. The function `erf' is the error function (see \citealt{stats book}). We choose an adaptive prior such that $\lambda_{i}\sim \text{Normal}\left(\hat{\lambda}, \sigma_{\lambda}\right)$, where $\hat{\lambda}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\lambda}\sim\text{Exponential}\left(1\right)$. This model can be re-parametrized following $q_1$, $q_2$ and $q_3$, and set the rest of the prior distributions as above (see Supplementary Table 2; Code Availability section). In this case, the Gaussian distribution is a special case of Eq.~(\ref{eq:skewed-response}) when $\lambda_{i}=0$ \citep{stats book}.

One way to test the extend to which species' physiological traits inform the skewness of the distribution of any species $i$ is by characterizing $\lambda_{i}$ as a Gaussian process. That is, we can consider the prior distribution for $\lambda_{i}$ as a multivariate normal with a variance covariance matrix $\Sigma^{\lambda}$. As described before, this variance-covariance matrix can be estimated by the model following Eq.~(\ref{eq:covariance-baseline}), and its structure can shed light on how informative the physiological traits are.

%\subsubsection*{Fat-tail and skewed response curve}
% Skewed generalized t distribution
% https://en.wikipedia.org/wiki/Skewed_generalized_t_distribution
% mcdonaldPartiallyAdaptiveRobust2016

\section*{Results}
% We first use the baseline model to understand how species are distributed along environmental gradient. As expected alpha declines with altitude, implying that the overall...  study the relationship between mean and variance. We found positive relationship... Hypothesis! Similar results were found using binary data. Different species show different correlaitons, invasive present less prevasive... When adding additional covariate, such the second axis of our PCA, this relationship vanishes... (see Supplementary)

%Other interesting comparisons... The value of adding plant physiology as prior information for both the mean and standard deviation of the distributions. Informative in both case, specially for... Capturing some of bla bla bla.

%Then, we stud the tails of distributions by modifying the baseline model to account for different properties. Kurtois. Studying the parameter... Similar values were found for binary data and the model did not outperform the baseline model

% Next we studied the skewness of distributions, outpreforming the other models. When studying the parameter, we found that the mean skewness of species distributions declined... Hypothesis! When accounting for plant physiology as priors for the skewness of distribution, we found this to be ... Figure and Hypothesis. The model also showcased the Rapopor's hypothesis

\section*{Discussion}
% Summary of results
% Proper test of Rapopor's hypothesis. Different species follow different biogeographical patterns.
% Proper test of skewed towards high altitude. Is species' physiology informative to explain the pattern? 
% What is the true shape of distributions? Missing bimodal curves
%

%%%%%%%%%%%%%%%%%%
%%% Model runs
%%%%%%%%%%%%%%%%%%
% Finish categorical 1d for both skewed and normal (2 weeks)
% Test for heavy tails 1 binomial and categorical (0.5 week)
% Test for heavy tails 2 binomial and categorical (0.5 week)
% Categorical 1d for heavy tails 1 (1 weeks)
% Categorical 1d for heavy tails 2 (1 weeks)
% Binomial 1d for heavy tails 1 (0.7 weeks)
% Binomial 1d for heavy tails 2 (0.7 weeks)
% Binomial 1d for skewed (0.7 weeks)
% Binomial 1d normal two covariances (1 week)
% Categorical 1d for skewed with physiology in lambda (1 week)
% Additional tests for Binomial 1d skewed and normal (1 week)

% With both heavy-tails ~ 10 weeks
% With only one heavy-tail ~ 8 weeks

\begin{figure}[ht]
  \centering
    \vspace{0.5cm}
    \includegraphics[width=0.9\textwidth]{figures/figure1}
    	  \vspace{0.3cm}
	   \caption{Relationship between mean and variance of species' distributions. These are the results for the main axis of variation for the climatic data (results for the second axis of variation presented in the Supplementary Fig.~2).}
      \label{fig:correlation}
\end{figure}
